Viscous Transport in Eroding Porous Media

### 2019-08-05 docs/transport/erosion_transport.tex
worked on tortuosity and anomalous diffusion sections

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Diffusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The dynamics of fluid in porous media is often characterized in terms of
anomalous diffusion~\cite{kla-rad-sok2008, den-cor-sch-ber2004}, and the
anomalous diffusion rate depends on the properties of the geometry.  The
anomalous diffusion depends on the porosity and
permeability~\cite{koc-bra1988}, but the distribution of the bodies also
affects the anomalous diffusion rates.  Since erosion causes certain
structure in porous media, calculating the anomalous diffusion rates in
these geometries is important.

One way to define the anomalous diffusion rate is by using tracer
trajectories that are governed by equation~\eqref{eqn:trajectories}.
Given a set of trajectories, we define $\lambda_j(t)$ to be the distance
travelled by the particle
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Having formed $N_p$ trajectories, the first and second ensemble moments
are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
At early times, the particles have not explored much of the geometry,
and we expect a ballistic motion $\sigma_\lambda^2 \sim t^2$.  However,
as the particles pass the grains, their trajectories are altered, and we
expect that $\sigma_\lambda^2 \sim t^\alpha$, with $\alpha < 1$, which
would indicate that the flow is sub-diffusive.  With respect to erosion,
as the bodies erode, we expect that the channelization of the geometry
results in more trajectories undergoing a ballistic motion, and this
would indicate a larger anomalous diffusion rate.

Towards the end of the erosion simulation, the porosity is low, and we
expect that the ensemble moments behave similarly to transport in a
pipe.  In a grainless geometry, the trajectory originating at
$(s_1(0),s_2(0))$ is
\begin{align}
  s_1(t) &= (1-s_2(0)^2)t + s_1(0), \\
  s_2(t) &= s_2(0).
\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry quantifies the average amount of turning
made by a streamline.  The choice of how to average is not universal,
and Duda et al.~\cite{dud-koz-mat2011} summarize several options for
averaging.  One particular choice, which we shall adopt, allows us to
compute the tortuosity in terms of trajectories, or by using an area
integral over the fluid geometry.  Another advantage of this choice is
that it is not affected by regions of recirculation.  Mathematically,
the definition of tortuosity that we use is
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is an cross-section perpendicular to the $x$ axis with length
$d$, $\lambda(y)$ is the length of the streamline that is initialized at
the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of the
velocity at the $y$-coordinate along $S$.  Since the flow is laminar in
the absence of grains, the tortuosity of a straight channel is $T=1$.
Following~\cite{dud-koz-mat2011}, two parallel cross-sections are
chosen, and the trajectories that pass between them are formed.  In our
work, we initialize the trajectories at $x=-1$ and terminate them at
$x=1$.

Under the assumptions that the flow is incompressible and not reentrant,
meaning that all streamlines connect the two cross-sections, the
tortuosity can be equivalently measured as~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy}.
  \label{eqn:tortuosity2}
\end{align}
Recirculation zones are possible in viscous fluids, but they only
constitute a small fraction of the total porous media, and in our
examples, their velocities are very small.  Therefore, the tortuosity
defined in equation~\eqref{eqn:tortuosity2} is an upper bound for the
definition in equation~\eqref{eqn:tortuosity1}, but for the geometries
we consider, we will show that their difference is negligible (see
Figure~\ref{fig:Eroding20tort_all}).  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories, we use this definition for the
majority of the tortuosity calculations.

The first method to numerically compute the tortuosity uses $N_p$
trajectories that are initially at $(-1,y_i)$, $i=1,\ldots,N_p$, and
stopped when their $x$ coordinate is 1.  We choose $y_i$ so that the
trajectories are initially uniformly spaced in $(-1,1)$.  Having
computed these trajectories, the tortuosity is approximated as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y_i}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.  Alternatively, to compute the tortuosity in terms of
the Eulerian velocity field, we first compute the velocity at $(x_i,y_j)
= (-1 + i\Delta x, -1 + j\Delta y)$, $i,j=1,\ldots,N$, where $\Delta x =
\Delta y = 2/N$, and the velocity at points inside an eroding body are
assigned a value of 0.  Then, the tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(x_i,y_j)\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u_x(x_i,y_j) \Delta x \Delta y}.
\end{align}

There have been efforts to relate the tortuosity to the
porosity~\cite{matyka2008tortuosity}, but these results can not be
universal since the porosity which depends on only a single number, the
tortuosity is affected by the distribution of pores within the porous
network.  In addition to the hydraulic tortuosity, a local tortuosity
can be used to analyze the local behavior of a streamline near a
body~\cite{matyka2008tortuosity}.  The local tortuosity of a streamline
initialized at $y$ along $S$ is
\begin{align}
  \tau(y) = \frac{\lambda(y)}{d}.
  \label{eqn:localTort}
\end{align}
\todo[inline]{Need to flesh this out more}
\todo[inline]{Talk about the different relationships between tortuosity
and porosity from the literature}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Diffusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The dynamics of fluid in porous media is often characterized in terms of
anomalous diffusion~\cite{kla-rad-sok2008, den-cor-sch-ber2004}.  The
anomalous diffusion depends on the porosity and
permeability~\cite{koc-bra1988}, but the distribution of the bodies also
affects the anomalous diffusion rates.  Since erosion causes certain
structure in porous media, calculating the anomalous diffusion rates in
these geometries is important.

We calculate the anomalous diffusion rates by analyzing the tracer
trajectories governed by equation~\eqref{eqn:trajectories} for
geometries that are formed by erosion.  Given a set of trajectories, we
define $\lambda_j(t)$ to be the arclength of the trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
Then, particle dispersion is characterized as $\sqrt{\sigma_\lambda^2}$.
At early times, the particles have not explored much of the geometry,
and we expect a ballistic motion $\sigma_\lambda^2 \sim t^2$.  However,
as the particles pass the grains, their trajectories are altered, and we
expect that $\sigma_\lambda^2 \sim t^\alpha$, with $\alpha < 1$,
indicating that the flow is sub-diffusive.  With respect to erosion, as
the bodies erode, we expect that the channelization of the geometry
results in more trajectories undergoing a ballistic motion, and this
would indicate decrease particle dispersion, meaning a larger anomalous
diffusion rate.

At early times, the geometry has not been altered by erosion, and we
expect large amounts of dispersion.  However, as the bodies erode, the
porosity increases and the trajectories behave more closely to laminar
trajectories in an open channel.  To be able to compare the dispersion
of an eroded geometry with that of an open channel, we consider
trajectories in an open channel.  The trajectory originating at
$(s_1(0),s_2(0))$ is
\begin{align}
  s_1(t) &= (1-s_2(0)^2)t + s_1(0), \\
  s_2(t) &= s_2(0).
\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}
In Section~\ref{sec:results}, we compare these moments in an open
channel with the trajectories in an eroding geometry.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry, denoted by $T \geq 1$, is a dimensionless
number that quantifies the average amount of turning made by
trajectories passing through the geometry.  The local tortuosity
\begin{align}
  \tau(y) = \frac{\lambda(y)}{d}.
  \label{eqn:localTort}
\end{align}
considers the length of a single trajectory, $\lambda$ whose initial
$y$-coordinate is a value $y$ on a cross-section of length $d$.  Taking
the average over all points on the cross-section defines the hydraulic
tortuosity
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is a cross-section perpendicular to the $x$-axis with length
$d$, $\lambda(y)$ is the length of the trajectory that is initialized at
the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of the
velocity at the $y$-coordinate along $S$.  Since the flow is laminar in
the absence of grains, the tortuosity of a straight channel is $T=1$.
The choice for the trajectory with length $\lambda(y)$ is not universal,
and Duda et al.~\cite{dud-koz-mat2011} summarize several options. One
choice, which we adopt, considers trajectories that originate at $S$ and
terminate at a parallel cross-section.  We use the cross-sections $x=-1$
and $x=1$.  This choice of averaging allows us to also compute the
tortuosity in terms of an area integral.  Under the assumptions that the
flow is incompressible and not reentrant, meaning that all streamlines
connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent to~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\cite{hig1985}, but they only constitute a small fraction of the
total porous media, and in our examples, their velocities are very
small.  Therefore, the tortuosity defined in
equation~\eqref{eqn:tortuosity2} is a tight upper bound for the
definition in equation~\eqref{eqn:tortuosity1}.  For the geometries we
consider, we will show that their difference is negligible (see
Figure~\ref{fig:Eroding20tort_all}).  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories, we use this definition for the majority
of the tortuosity calculations.

The first method to numerically compute the tortuosity uses $N_p$
trajectories that are initialized at $(-1,y_i)$, $i=1,\ldots,N_p$, and
stopped when their $x$ coordinate is 1.  We choose $y_i$ so that the
trajectories are initially uniformly spaced in $(-1,1)$.  Having
computed these trajectories, the tortuosity is approximated as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.  Alternatively, to compute the tortuosity in terms of
the Eulerian velocity field, we first compute the velocity at $(x_i,y_j)
= (-1 + i\Delta x, -1 + j\Delta y)$, $i,j=1,\ldots,N$, where $\Delta x =
\Delta y = 2/N$, and the velocity at points inside an eroding body are
assigned a value of 0.  Then, the tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(x_i,y_j)\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u_x(x_i,y_j) \Delta x \Delta y}.
\end{align}

There have been efforts to relate the tortuosity to the
porosity~\cite{matyka2008tortuosity}.  For example, Matyka et
al.~\cite{matyka2008tortuosity} propose the models
\begin{align}
  T(\phi) &= \phi^{-p}, \\
  T(\phi) &= 1-p \log \phi, \\
  T(\phi) &= 1+p (1-\phi), \\
  T(\phi) &= (1+p (1-\phi))^2, 
\end{align}
where $p>0$ is a fitting parameter.  In Section~\ref{sec:results}, we
compare these four models with the tortuosity of eroding geometries.
